\documentclass[msc,british,bibkey]{ThesisPUC_uk}

%---------- Math ----------%
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{bbm}
%---------- Floting ----------%
\usepackage{float}
%---------- References ----------%
\usepackage{natbib}
%---------- Algorithm ----------%
\usepackage{algorithm}
\usepackage{algorithmic}
%---------- Tables ----------%
\usepackage{rotating}
\usepackage{tabularx}
\usepackage{multicol}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{subcaption}
\usepackage{diagbox}
%---------- Others ----------%
\usepackage{hyperref}
\usepackage{bold-extra}
\usepackage{graphicx}

%\colorlet{dark}{red!85!blue!60!black}

%---------- Cover ----------%
\author{\mbox{Felipe} \mbox{de} \mbox{Albuquerque} \mbox{Mello} \mbox{Pereira}}
\authorR{\mbox{Pereira}, \mbox{Felipe} \mbox{de} \mbox{Albuquerque} \mbox{Mello}}
\advisor{\mbox{Eduardo} \mbox{Sany} \mbox{Laber}}
\advisorR{\mbox{Laber}, \mbox{Eduardo} \mbox{Sany}}
\title{Binary Splitting Criteria for Large Categorical Attributes in Decision Trees}
\day{28$^{th}$} \month{February} \myyear{2018}

\city{Rio de Janeiro}
\CDD{004}
\department{Inform\'atica}
\departmentbr{Inform\'atica}
\program{Inform\'atica}
\programbr{Inform\'atica}
\school{Centro T\'{e}cnico Cient\'{\i}fico}
\university{Pontif\'{\i}cia Universidade Cat\'{o}lica do Rio de Janeiro}
\uni{PUC-Rio}

%---------- Jury ----------%

\jury{
  \jurymember{H\'elio C\^ortes Vieira Lopes}{Departamento de Inform\'atica --- PUC-Rio}
  \jurymember{Marco Serpa Molinaro}{Departamento de Inform\'atica --- PUC-Rio}
  \jurymember{M\'arcio da Silveira Carvalho}{Vice Dean of Graduate Studies\\ Centro T\'ecnico Cient\'ifico da PUC-Rio}
}

%---------- Front letters ----------%
\resume
{
Bachelor's in Electrical Engineering and Pure Mathematics at the Pontif\'icia Universidade Cat\'olica do Rio de Janeiro (2010 and 2011). Masters' in Mathematics at the Pontif\'icia Universidade Cat\'olica do Rio de Janeiro (2013).
}

\acknowledgment
{
TODO: acknowledgment.

Thanks to CNPq for the conceded scholarship during my Masters.
}

\keywords
{
  \key{Decision Trees; Max-cut Problem; Approximated Algorithms}
}

\abstract{
In this dissertation we proposed a framework for designing splitting criteria for handling multi-valued nominal attributes for decision trees. Criteria derived from our framework can be implemented to run in polynomial time in the number of classes and values, with theoretical guarantee of producing a split that is close to the optimal one. This is the only known criteria that have all these characteristics simultaneously. We also run multiple experiments to evaluate its running times and accuracy in real datasets.
}

\titulobr{Crit\'erios de Splits Bin\'arios para Atributos Categóricos Grandes em \'Arvores de Decis\~ao}
\departamentobr{Inform\'atica}

\chavesbr
{
  \chave{\'Arvores de Decis\~ao; Problema de Corte M\'aximo; Algoritmos Aproximativos}
}

\resumobr
{
Nesta disserta\c{c}\~ao \'e apresentado um framework para desenvolver crit\'erios de split para lidar com atributos nominais multi-valorados em \'arvores de decis\~ao. Crit\'erios gerados por este framework podem ser implementados para rodar em tempo polinomial no n\'umero de classes e valores, com garantia te\'orica de produzir um split pr\'oximo do \'otimo. Este \'e o \'unico crit\'erio conhecido que possui ambas caracter\'isticas simultaneamente. Tamb\'em s\~ao realizados m\'ultiplos experimentos para avaliar seu tempo de execu\c{c}\~ao e acur\'acia em datasets reais.
}

\tablesmode{figtab}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\newcommand{\remove}[1]{}
\newcommand{\dist}[3]{d(#1, #2)}
\newcommand{\distC}[2]{d(#1, #2)}
\newcommand{\OPT}[1]{\textrm{OPT}(#1)}
\newcommand{\OPTf}[2]{\textrm{OPT}(#1, #2)}
\newcommand{\Cf}[2]{\textrm{cost}(#1, #2)}
\newcommand{\C}[1]{\textrm{cost}(#1)}
\newcommand{\ans}[1]{\textbf{#1}}
\newcommand{\bl}{\textrm{blocked}}
\newcommand{\un}{\textrm{unassigned}}
\newcommand{\comments}[1]{}
\newcommand{\commento}[1]{\marginpar{\tiny \flushleft{#1}}}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}[section]

\input{Introduction}
\input{Background}
\input{Framework}
\input{ExperimentsSplits}
\input{ExperimentsDatasets}
\input{Conclusions}

\begin{raggedright}
	\bibliography{thesis}
\end{raggedright}

\end{document}